Using the R console as a calculator
========================================

* Some R data types
    * Use the `class()` function to show the data type
    * numeric
    * logical
    * character
    * vectors
    
```{r data types}
# numeric
class(1)
class(pi)

# complex
class(3+4i)

# logical
class(TRUE)
class(T)
class(FALSE)
class(F)

# character
class("R is cool!")

# vectors
xs <- c(1, pi, 3+4i)
class(xs)

# dates
Sys.Date() # today

# working with strings
help(cat)
cat("Hello, world")
dates <- seq.Date(Sys.Date(), by=7, length.out=4)

# suppose you want to create a set of filenames with a date in the filename
help(paste)
paste("experiment-", dates, ".fcs", sep="")
```
      
* Variables and the assignment operator

```{r assignment}
# we "store" stuff for later use with the <- assignment operator
a <- 3
b <- 7
a*b

# variable names can include '.'
this.is.a.long.name <- a*b
this.is.a.long.name
```

* Type `??operators` in the console and explore the use of Arithmetic, Colon, Relational and Logical operators
```{r operators}
# Arithmetic
5+(2*3)^7
5 %% 2 # modulus operator - remainder after division

# Logical
TRUE && FALSE # logical AND
TRUE || FALSE # logical OR
!TRUE # negation

# Relational
3 < 4
3 > 4
3 == 4
3!= 4

# Colon (from:to)
10:20

# a generalization of the from:to colon operator is the seq() function
seq(20, 2, by=-2)
seq(10, by=1, length.out=5)
seq(0, 2*pi, pi/4)
```

* Type `?Trig` to see what trigonometric functions are available
* Type ?`?log` to see how to do logarithms and exponentials.

A simple example combining trig and exponential functions tests Euler's famous formula $e^{i\theta} = cos(\theta) + i \sin(\theta$
```{r trig transcendentals}
theta <- seq(0, 2*pi, pi/4)
exp(theta*1i) == cos(theta) + 1i*sin(theta)
```

* Type `demo(graphics)` to see some of R's basic graphics capabilities
```{r graphics}
demo(graphics)
```



Exercises
----------------------------------------

* Use the combine `c()` function to make a vector of the numbers 23, 47, 99, and find the mean, median and variance of those numbers. Remember to assign the vector to a variable.
* Create a vector from 0 to 2\\( \pi \\) of length 100 and assign to a variable \\( x \\)
* Plot sin(x)
* Customize the plot into a smooth red line instead of dots
