Tour of RStudio
========================================

Web references
--------------------------------------------------------

* [R manuals](http://cran.r-project.org/manuals.html)
* [RStudio docs](http://www.rstudio.com/ide/docs/)
* [R search engine](http://www.rseek.org/)

Installation
----------------------------------------

Follow instructions at http://www.rstudio.com/ide/download/desktop, making sure to install R first from http://cran.rstudio.com/.

Anatomy of RStudio
----------------------------------------

### Menu
* File
    * Create, open and save R scripts
* Session
    * Change the current working directory if you are not where you expect to be
* Project
    * Create or open R projects
* Tools
    * Load data sets and install new R packages

### Panels
* Editor
* Console
    * Prompt (>)
    * Getting help with (?) or help()
```{r}
help(seq) # shortcut is ?seq
```
    
    * Searching by keywords (??) or help.search()

```{r}
help.search("chisquare") # shortcut is ??chisquare
```
    * Tab completion - e.g. seq()
    * History with up and down arrows
    * History completion with C-up
    * Incomplete commands show up as (+)
    * Escaping from '+' prompt with ESC key
* Workspace | History
    * Workspace shows user-defined variables that R currently knows about
    * History shows previous commands
        * Re-running a command with "To Console" (or Enter)
        * Assembling a script with "To Source" (or Shift-Enter)
* Files | Plots | Packages | Help
    * Files shows the contents of the current working directory
    * Plots
        * Export
    * Packages manages the installed code base
    * Help provides hyperlinked documentation
        * Search text box
        * Drop down menu for Help history
        * Forward and Back buttons


Exercises
----------------------------------------

* Create a new project called RWS1
* Install the knitr and plyr packages
* Get help on the seq() function and use it to generate the sequence 2.0, 2.5, 3.0, 3.5 4.0
* Find help for all the R functions that contain the string "seq"
* Create a new R script in the Editor by using File | New
* Use the History panel and "To Source" to send the sequence generation command to the newly created R script
* Run the command by clicking on the "Source" button in the Editor panel