Lab 5: Data summary and visualization with R
========================================================
Datasets
--------------------------------------------------------

### Built-in datasets in R
```{r fig.width=4, fig.height=4}
data() # all available datasets
data(package = 'lattice') # all data sets in a package
```
```{r fig.width=4, fig.height=4}
data(barley, package = 'lattice') # read in the dataset without loading the package
help(barley, package = 'lattice') # get information about the dataset without loading the package
```
```{r fig.width=4, fig.height=4}
# same as
library(lattice) # load the package
data(barley) # read in the dataset
help(barley) # get information about the dataset
```
```{r fig.width=4, fig.height=4}
head(barley)
```

### External datasets 
* http://exploringdata.net/datasets.htm

* http://lib.stat.cmu.edu/DASL/ 

* http://www.statsci.org/datasets.html

  and more...

A dataset from http://lib.stat.cmu.edu/DASL/ :
```
Story Names: Scents and Learning

Description: Data on the time subjects required to complete a pencil and paper maze when they were smelling a floral scent and when they were not.

Number of cases: 21

Variable Names:
ID:
Sex: M=male, F=female
Age: Age in years
Smoker: Y if subject smoked, N if did not
Opinion: "pos" if subject found the odor inherently positive, "indiff" if indifferent, "neg" if inherently negative
Order: 1 if did unscented trials first, 2 if did scented trials first
UTrial1: length of time required for first unscented trial
UTrial2: length of time required for second unscented trial
UTrial3: length of time required for third unscented trial
STrial1: length of time required for first scented trial
STrial2: length of time required for second scented trial
STrial3: length of time required for third scented trial
```
```{r fig.width=4, fig.height=4}
sc = read.table("scents.txt", head = T) # read in the data
names(sc)
head(sc)
# Access a variable:
sc$Age
# Access a variable by its name:
Age
attach(sc) # to access the variable names 
Age
detach(sc) # clean up
```

Summary statistics and graphical displays
--------------------------------------------------------
This simple command gives a (univariate) summary of each column:
```{r fig.width=4, fig.height=4}
summary(sc) # gives mean and five number summary for numeric type, and counts for charactor type
```

### Univariate data I - categorical
#### Factors
* In the above dataset, there are five categorical variables, 'ID', 'Sex', 'Smoker', 'Opinion', and 'Order'. 
* A categorical variable is also called a factor. 
```{r}
# A character vector is automatically treated as a factor: 
is.factor(sc$Sex)
sc$Order
is.factor(sc$Order) # 'Order' is coded as numeric but in fact a categorical variable! 
```

```{r fig.width=4, fig.height=4}
sc$Order = factor(sc$Order) # now treated as a categorical variable
sc$Order # levels are printed; levels are numeric
```
#### Frequency tables 
```{r fig.width=4, fig.height=4}
table(sc$Order)
# same as
summary(sc$Order)
```
#### bar plots
```{r fig.width=4, fig.height=4}
barplot(sc$Smoker) # using raw data?
barplot(table(sc$Smoker)) # using summarized data!
barplot(table(sc$Smoker), col = c("light blue", "pink")) # change color
barplot(table(sc$Smoker)/length(sc$Smoker)) # proportions
table(sc$Smoker)/length(sc$Smoker) # proportions
```
#### Pie charts
```{r fig.width=4, fig.height=4}
(tb <- table(sc$Opinion)) # make assignment as well as print the object
pie(tb)
names(tb) = c("Indifferent", "Negative", "Positive") # new labels for levels
pie(tb) # now with new labels
```

### Univariate data II - numerical
#### Five number summary
```{r}
fivenum(sc$Age) # min, lower hinge, dedian, upper hinge, max
summary(sc$Age) # five number summary and mean
```
#### Stem-and-leaf plots
```{r}
stem(sc$UTrial1)
stem(sc$UTrial1, scale = 2) # a longer plot
```
#### Histograms
```{r fig.width=4, fig.height=4}
hist(sc$STrial1)
hist(sc$STrial1, prob = T) # proportions; area under the histogram equals 1
rug(sc$STrial1) # add tick marks
lines(density(sc$STrial1), lty = 2) # add a dotted density curve 
```
#### Box plots
```{r fig.width=4, fig.height=4}
boxplot(sc$STrial2)
boxplot(sc$STrial2, main = "Box plot for S-Trial 2", ylab = "Time (second)") # Add title and y-axis label
rug(sc$STrial2, side = 2, col = "red") # add tick marks
```

### Bivariate data I - categorical vs. categorical
#### Contingency tables
```{r fig.width=4, fig.height=4}
table(sc$Smoker, sc$Sex)
table(Smoker = sc$Smoker, Sex = sc$Sex) # prints names as given
```
#### bar plots
```{r fig.width=4, fig.height=4}
barplot(table(sc$Smoker, sc$Sex))
barplot(table(sc$Sex, sc$Smoker))
barplot(table(sc$Sex, sc$Smoker), beside = T) # besides; default is stacked
barplot(table(sc$Sex, sc$Smoker), beside = T, legend.text = T) # add a legend
```
### Bivariate data II - categorical vs. numerical
#### Box plots 
```{r fig.width=4, fig.height=4}
boxplot(sc$UTrial1 ~ sc$Sex) # a side by side box plot
```
```{r fig.width=4, fig.height=4}
# same as above except for x-axis labels
boxplot(sc$UTrial1[sc$Sex == "F"], sc$UTrial1[sc$Sex == "M"]) 
```
```{r fig.width=4, fig.height=4}
# change axis labels
boxplot(sc$UTrial1[sc$Sex == "F"], sc$UTrial1[sc$Sex == "M"], xaxt = "n") # no x-axis
axis(1, at = 1:2, labels = c("F", "M")) # new x-axis
rug(sc$UTrial1[sc$Sex == "F"], side = 2) # add a rug
rug(sc$UTrial1[sc$Sex == "M"], side = 4) # add a rug
```
#### Stripcharts (dot plots)
```{r fig.width=4, fig.height=4}
stripchart(sc$UTrial1 ~ sc$Sex)
```
```{r fig.width=4, fig.height=4}
# Add a stripchart to a box plot:
boxplot(sc$UTrial1 ~ sc$Sex)
stripchart(sc$UTrial1 ~ sc$Sex, vertical = T, col = "red", pch = "x", add = T)
```
### Bivariate data III - numerical vs. numerical
#### Correlation coefficients
Strengh of the 'linear' relationship between two numerical variables
```{r fig.width=4, fig.height=4}
cor(sc$UTrial1, sc$STrial1)
cor(sc[, 7:12]) # correlation matrix 
```
#### Scatter Plots
```{r fig.width=4, fig.height=4}
plot(sc$UTrial1, sc$UTrial2)
abline(lm(sc$UTrial2 ~ sc$UTrial1)) # add the regression line of UTrial2 on UTrial1 (y on x)
```
```{r fig.width=4, fig.height=4}
plot(sc$UTrial1, sc$UTrial2, type = "n") # no points or lines
points(sc$UTrial1, sc$UTrial2, pch = 19, cex = 2) # plot points
plot(sc$UTrial1, sc$UTrial2, type = "n") # no points or lines
text(sc$UTrial1, sc$UTrial2, labels = sc$Smoker) # plot character strings
```
```{r fig.width=4, fig.height=4}
# All pairwise scatter plots:
plot(sc[, 7:12])
```

```{r fig.width=4, fig.height=4}
# 'plot' for numerical vs. categorical?
plot(sc$Sex, sc$UTrial1) 
```
```{r fig.width=4, fig.height=4}
# same as 
boxplot(sc$UTrial1 ~ sc$Sex)
```
```{r fig.width=4, fig.height=4}
# also same as this except for x, y axis labels
plot(sc$UTrial1 ~ sc$Sex)
```



R Graphics
--------------------------------------------------------
### High-level plotting functions
* 'plot' = scatter plot

* 'hist' = histogram

* 'boxplot' = box plot

* 'stripchart' = strip chart

* 'barplot' = bar plot

* 'stem' = stem-and-leaf plot

* 'curve' = curve corresponding to a function or an expression.

  
  and more ... 

```{r}
# Arguments of the 'plot' function:
help(plot)
```
```
* type: "p" = points; "l" = lines; "b" = both; "c" = empty points joined by lines; "o" = overplotted points and lines; "s" or "S" = stair steps; "h" = histogram-like vertical lines; finally, "n" does not produce any points or lines.

* xlim, ylim: x limits,  y limits

* main: a main title

* sub: a sub title

* xlab, ylab: a label for the x-, y- axis

* axes: a logical value indicating whether both axes should be drawn on the plot. Use "xaxt" or "yaxt" to suppress just one of the 

* xaxt: The x axis type. Specifying "n" suppresses plotting of the axis. The standard value is "s". any value other than "n" implies plotting.
```
  
### Lower-level plotting functions
To add further elements to the main plot
```{r}
help(plot.default)
help(par)
```

* 'axis' adds an axis on the specified side.

* 'title' adds titles (main, sub) and axis lables.

* 'lines' draws single line that connect the points at the specified coordinates.

* 'abline' draws single straight line with the specified slope and intercepts. 

* 'text' draws a sequence of strings at the specified coordinates.

* 'points' draws a sequence of points at the specified coordinates.

* 'legend' adds a legend.

* 'rug' adds tick marks on the specified side.


### Setting graphical parameters using 'par'
Permanant changes in graphical parameters for the current graphic device. 
```{r}
help(par)
```

Combining multiple plots:
```{r fig.width=4, fig.height=4}
par(mfrow = c(2,2)) # a 2x2 matrix of plots filled in by row
x = seq(-pi/4, pi/4, .01)
plot(x, sin(x), type = 'l')
plot(x, cos(x), type = 'l')
plot(x, tan(x), type = 'l')
```

Restoring original settings (from Quick-R http://www.statmethods.net/advgraphs/parameters.html): 
```{r eval = F}
par()              # view current settings
opar <- par()      # make a copy of current settings
par(col.lab="red") # red x and y labels 
hist(mtcars$mpg)   # create a plot with these new settings 
par(opar)          # restore original settings
```

Exercise
--------------------------------------------------------
* Read in the 'barley' dataset in the 'lattice' package. 

* Attach the data.

* Obtain univariate summary statistics for each variable. 

* Make a histogram for yield. 

* Make a contingency table for variety and site. 

* Using a side-by-side box plot, investigate the relationship between yield and variety. Repeat this for yield and site. 

* Enter in the following data. 

  x: 4  9  9 10  9  8  4  7  9  3
  
  y: 9 19 20 21 18 17  9 15 20  7
  
* Make a scatter plot of x and y. 

* Add a straight line with slope = 2, y-intercept = 1 using one of the following. 

  * abline(a = 1, b = 2)
  
  * s = seq(3, 10, .1); lines(s, 2*s + 1)
  
  * curve(2*x + 1, 3, 10, add = T)

* Do not leave RStudio until you learn how to save your work in the next lab session. 
  