Lab 2: Motivating examples in R
========================================================

## Read a CSV data file into a data frame
```{r}
aids = read.csv("./lab4_data/aids.csv", check.names = F)
head(aids)
```
## Use basic R functions to summarize the data
### Five number summary
```{r}
data(mtcars) # built-in dataset
head(mtcars)
fivenum(mtcars$mpg)
```
### Contingency table
```{r}
table(gear = mtcars$gear, carb = mtcars$carb)
```
## Manipulate the data
### Reshape data
```{r}
data(PlantGrowth) # built-in dataset
head(PlantGrowth) # 30 rows
unstack(PlantGrowth) # 10 rows: reshaped from long to wide format
```
### Subset data
```{r}
subset(PlantGrowth, PlantGrowth$weight > 5.5)
```
## Perform simple analysis
### ANOVA
```{r}
summary(aov(PlantGrowth$weight ~ PlantGrowth$group))
```
### Linear regression
```{r}
myreg = lm(mtcars$mpg ~ mtcars$wt)
summary(myreg)
```
## Generate graphs of the analysis
### Scatter plot with regression line
```{r fig.width=5, fig.height=5}
plot(mtcars$wt, mtcars$mpg, xlab = "weight", ylab = "mpg", main = "Scatter plot for weight and mpg") 
abline(myreg) # add regression line
```
## Save results of analysis and make hard copies of graphics 
### Saving graphs 
* Directly save to a file without plotting on an active window:
```{r eval=FALSE}
# open a new graphic device:
pdf("myfigure1.pdf") # other formats: jpeg (.jpg), png (.png), postscript (.ps), bmp (.bmp)
plot(mtcars$wt, mtcars$mpg, xlab = "weight", ylab = "mpg", main = "Scatter plot for weight and mpg") 
abline(myreg) # add regression line
dev.off()
```
* Graphics already plotted on an active window can be saved using the 'Export' command. 

### Saving tables
* Directly save to a file without printing the object: 
```{r eval=FALSE}
result.tb = table(mtcars$gear, mtcars$carb)
write.table(result.tb, "mytable.txt", quote = F) # write a text file 
write.csv(result.tb, "mytable.csv") # write a csv file
```
* You can also copy the printed table from the R console and paste to your text editor to edit and/or save. 

### Saving workspace
```{r eval=FALSE}
save.image() # save the current workspace
```
```{r eval=FALSE}
load(".RData") # load the saved values
```
### Printing graphics
* Open saved graphics and print them out.

## Create an HTML report with knitr
This HTML document was generated in RStudio using knitr!