Lab 6: Reproducible analysis with R
========================================================
Could someone else produce the exact same results on the same data? 

## Basic: preserve input files and save results
### Preserving input files
http://www.dataone.org/best-practices/preserve-information-keep-your-raw-data-raw

Keep the raw data raw.
### Saving graphs 
Directly save to a file without plotting on an active window:
```{r eval=FALSE}
data(mtcars)
names(mtcars)

# open a new graphic device:
pdf("myfigure1.pdf") # other formats: jpeg (.jpg), png (.png), postscript (.ps), bmp (.bmp)
plot(mtcars$wt, mtcars$mpg, xlab = "weight", ylab = "mpg", main = "Scatter plot for weight and mpg") 
abline(myreg) # add regression line
dev.off()
```
```{r eval=FALSE}
pdf("c:/myfigures/myfigure1.jpg", height = 5, width = 5) # save in the specified folder, not the working directory, with specified size
plot(mtcars$wt, mtcars$mpg, xlab = "weight", ylab = "mpg", main = "Scatter plot for weight and mpg") 
abline(myreg) # add regression line
dev.off()
```
Graphics already plotted on an active window can be saved using the 'Export' command. 

### Saving tables
Directly save to a file without printing the object: 
```{r eval=FALSE}
result.tb = table(mtcars$gear, mtcars$carb)
write.table(result.tb, "mytable.txt", quote = F) # write a text file 
write.csv(result.tb, "mytable.csv") # write a csv file
```
You can also copy the printed table from the R console and paste to your text editor to edit and/or save. 

### Saving workspace
```{r eval=FALSE}
save.image() # save the current workspace
```
```{r eval=FALSE}
load(".RData") # load the saved values
```
### Saving R command history 
* Can be saved using the 'To Source' command. 

* Then edit (delete unwanted commands) and add comments. 

## Writing R scripts 
* R commands + comments = R script

* A well documented script should contain detailed comments. 

* Can be reviewd, edited or rerun (by the 'source' command) later on. 

## Using R Markdown and knirt 
* Pretty HTML documentation

* Literate programming  

* Reproducible template for basic tasks in R

* To reproduce this document, 
  1. Open the R Markdown file 'Lab6.Rmd' in RStudio.
  2. Click the 'Knit HTML' command on the File menu or on the source editor toolbar. 

* Using R Markdown with RStudio
  http://www.rstudio.com/ide/docs/authoring/using_markdown

* Creating Notebooks from R Scripts
  http://www.rstudio.com/ide/docs/authoring/markdown_notebooks


Exercise
--------------------------------------------------------

* Save your graphs, history, and workspace from the previous lab exercise. 

* Create a new R Markdown (template) and click the 'knitr' button to generate an HTML document. 

* Create a new R Script including the following lines (including comments). 
  ```
  # Generating a vector of even numbers between 0 and 10 
  # 1. using 'seq':
  seq(0, 10, by = 2)
  # 2. using scalar multiplication of a vector:
  0:5*2
  ```

* Then generate an R Markdown file and a HTML document using the 'Compile Notebook' command. 

