Lab 8: Linear regression (continuous outcomes)
========================================================
Simple linear regression
--------------------------------------------------------
```{r fig.width=4, fig.height=4}
Predictor = c(2, 4, 6, 10, 18, 19, 21, 22, 24, 40)
Outcome = c(16, 18, 20, 22, 24, 25, 28, 29, 33, 50)
mylm = lm(Outcome ~ Predictor)
mylm # printing an lm object: not much to see
slm = summary(mylm)
slm # printing the summary of an lm object
# What are in the summary of an lm object?
names(slm)
# coefficient estimates
mycoef = slm$coef 
mycoef
# or simply
coef(mylm) # a (named) vector
round(coef(mylm), 2) # rounding
# 'mycoef' is in a matrix form, so try this for column names
names(mycoef) 
is.data.frame(mycoef) # 'mycoef' is a matrix, but not a data.frame
# 'names' works with data.frame, but not with general matrices To get column names of general matrices, use
colnames(mycoef)
# standard errors for coefficient estimates
mycoef[, 2]
# same as
mycoef[, "Std. Error"]
# also same as
as.data.frame(mycoef)$St # abbreviation does work here
# confidence intervals for coefficient estimates
confint(mylm)
confint(mylm, level = .9)
confint(mylm, "Predictor", .9)
# residual standard error
slm$sigma
# What are in an lm object?
names(mylm)
# coefficient estimates again
mylm$coef # abbreviation works
# fitted values
mylm$fitt 
# same as
fitted(mylm) # no abbreviaion allowed for function names
# residulas
mylm$res
# same as
residuals(mylm)
# also same as
resid(mylm)
# also same as
slm$res
# also same as 
Outcome - mylm$fitt
# predictiction
morePredictor = c(15.5, 17.5)
X = data.frame(Predictor = morePredictor)
predict(mylm, newdata = X)
# same as
mycoef[1] + mycoef[2] * morePredictor
```

Multivariate linear regression
--------------------------------------------------------
```{r fig.width=4, fig.height=4}
data(environmental, package = 'lattice')
head(environmental)
plot(environmental)
lmozone = lm(ozone ~ radiation + temperature + wind, data = environmental)
summary(lmozone)
```
The fitted regression equation can be written as
```
ozone = `r coef(lmozone)[1]` + `r coef(lmozone)[2]` * radiation + `r coef(lmozone)[3]` * temperature + `r coef(lmozone)[4]` * wind  
```
Prediction of ozone for new covariate values can be done with this equation or the 'predict' command. 

Exercise
--------------------------------------------------------
* For the above Predictor and Outcome data, create a scatter plot for Outcome versus Predictor and add the fitted regression line. 

* Plot residuals aginst fitted values. This is called a residual plot. Add a horizontal line at 0. 

* For the built-in dataset 'women', fit a linear regression model for weight against height. 

* Find fitted values and the $R^2$ statistic.

* Create a scatter plot for weight against height. Add the fitted regression line. 

* Predict weights of three women with heights being 65, 59, 62, respectively, in inches. 
