Lab 5: Data manipulation and exploratory data analysis
========================================================

## Names
### preserving original names
```{r}
# default: syntactically invalid variable names are adjusted 
scent1 = read.table('scent_orig.txt', sep = '\t', head = T)
head(scent1)
# with option check.names = FALSE
scent2 = read.table('scent_orig.txt', sep = '\t', head = T, check.names = F)
head(scent2)
````
### retrieving or setting names 
* `dimnames`, `colnames`, `rownames`: work with general matrices

* `names`, `row.names`: methods for data frames

* `names` also works with vectors

```{r} 
### 1. retrieving names:

# row and column names
dimnames(scent2)
# row names
dimnames(scent2)[[1]] 
# same as 
rownames(scent2)
# also same as
row.names(scent2) # only for data frames
# column names
dimnames(scent2)[[2]]
# same as
colnames(scent2)
# also same as 
names(scent2) # only for data frames and vectors
### 2. setting row or column names: assign a character vector

rownames(scent2) = paste("row", 1:nrow(scent2), sep = ".")
rownames(scent2)
```

## Missing values
### checking missingness
```{r}
# a data frame containing missing values
dat = data.frame(v1 = c(1, 2, 3, NA), v2 = c(5, NaN, 7, 8), v3 = 9:12)
dat
# check missingness
is.na(dat) # includes NaN!
# TRUE indices only
which(is.na(dat))
# return array indices
which(is.na(dat), arr.ind = T)
# any missing?
any(is.na(dat))
# number of missing values
sum(is.na(dat))
# same as
length(which(is.na(dat)))
```
### setting NA to a value
```{r}
dat[is.na(dat)] = 99
dat
```
### setting elements to NA
```{r}
# a matrix without NA
dat2 = matrix(1:12, nrow = 4)
dat2
is.na(dat2) = c(4, 6)
dat2
```
### removing NA
```{r}
# complete rows only
complete.cases(dat2) 
dat2[complete.cases(dat2), ]
# same as
na.omit(dat2)
# complete columns only
complete.cases(t(dat2))
dat2[, complete.cases(t(dat2))]
# same as above, retaining the number of dimensions
dat2[, complete.cases(t(dat2)), drop = F]
# error with NA
mean(dat2)
# with NA removed
mean(dat2, na.rm = T)
```
## Sorting
* `sort` to sort a vector

* `order` to sort data frames, etc. It returns the permutation which orders the data. 

### sort
```{r}
x = sample(1:4, 10, replace = T)
x
# sort in ascending order
sort(x)
# sort in descending order
sort(x, decreasing = T)
```
### order
```{r}
y = sample(1:4, 10, replace = T)
cbind(x, y)
# permutation that sorts x
order(x)
# use y for breaking ties
order(x, y)
# minus sign for descending order
order(-x, y)
# sort data frame by cyl and then gear
mtcars.s = mtcars[order(-mtcars$cyl, -mtcars$gear), ]
head(mtcars.s)
# compare with the original data
head(mtcars)
```

## Subsetting
### Using indices
```{r}
# extracting rows
mtcars[mtcars$gear == 4 & mtcars$carb == 1, ]
# extracting columns
mtcars[1:5, c(2, 4)]
# same as 
mtcars[1:5, c("cyl", "hp")]
# also same as
mtcars[1:5, names(mtcars) %in% c("v1", "v2", "cyl", "hp", "v3")]
# use the minus sign for exclusion
mtcars[1:5, -c(2, 4)]
# but this does not work 
mtcars[1:5, -c("cyl", "hp")]
# instead, this works
mtcars[1:5, setdiff(names(mtcars), c("cyl", "hp"))]
```
### which
```{r}
which(mtcars$gear == 4 & mtcars $ carb == 1)
which(mtcars == 6)
which(mtcars == 6, arr.ind = T)
which.max(mtcars$mpg)
```
### subset
```{r}
subset(mtcars, mtcars$am == 1 & mtcars$cyl == 8)
subset(mtcars, mtcars$am == 1 & mtcars$cyl == 8, select = mpg:hp)
```

## Combining
* `rbind`, `cbind` to combine a seqence of vector, matrix or data frames by rows or columns

* `merge` to merge two data frames by common columns or row names

### By assignment
```{r}
datxy = data.frame(x, y)
datxy$z = datxy$x + datxy$y
datxy
```
### rbind, cbind
```{r}
# combine by rows
rbind(x, y)
# combine by columns
cbind(x, y)
# similar to 
data.frame(x, y)
# combine a data frame and a vector
col125.A = cbind(mtcars[, 1:2], mtcars[, 5])
col125.A[1:5, ]
# assign a name
col125.B = cbind(mtcars[, 1:2], drat = mtcars[, 5])
col125.B[1:5, ]
# data.frame also works
data.frame(mtcars[, 1:2], drat = mtcars[, 5])[1:5, ]
# same as 
col125.C = mtcars[, c(1:2, 5)]
col125.C[1:5, ]
```

### merge
```{r}
a = data.frame(c(list(names=c("Jack", "Jill", "Tom"), ages=c(10, 12, 14))))
a
b = data.frame(c(list(kids=c("Jack", "John", "Jill"), states=c("NC", "SC", "MA"))))
b
# merge two data frames
merge(a, b, by.x = "names", by.y = "kids")
# include all values in the first data frame
merge(a, b, by.x = "names", by.y = "kids", all.x = T)
```



## Reshaping
* wide format: repeated measurements in separate columns of the same record

* long format: repeated measurements in separate records

* `stack`: from wide to long

* `unstack`: from long to wide

* `reshape`

### stack, unstack
```{r}
data(Indometh)
head(Indometh)
# unstack to the wide format
unstack(Indometh, form = conc ~ time)
Ind.wide = unstack(Indometh, form = conc ~ Subject)
Ind.wide
# stack it back (into the long format) with the select option
stack(Ind.wide, select = -c(X1, X2, X3, X4))
```

### reshape
```{r}
wide = reshape(Indometh, v.names = "conc", idvar = "Subject", timevar = "time", direction = "wide")
wide
reshape(wide, direction = "long")[1:5, ]
reshape(wide, idvar = "Subject", varying = list(2:12), v.names = "conc", direction = "long")[1:5, ]
```

## Splitting
* `split` to divide data into groups 

* `cut` to categorize numeric values

* `strsplit` to split strings


## Applying functions
* `apply` applies a function to margins (rows or columns)

* `sapply` applies a function over a list (e.g., data frame) or vector

* `tapply` applies a function to a vector by group

* `aggregate` applies a function by group like `tapply`, over a data frame

* `by` applies a function to a data frame by group

* `rowSums`, `colSums`, `sweep`, `scale`

### apply
```{r}
mymat = matrix(rnorm(12), 3, 4)
mymat 
# row sums
apply(mymat, 1, sum)
#same as 
rowSums(mymat)
# column sums
apply(mymat, 2, sum)
#same as 
colSums(mymat)
```
### sapply
```{r}
# column sums
sapply(as.data.frame(mymat), sum)
# this is not what we want
sapply(mymat, sum)
```
### tapply
```{r}
# mean mpg by cyl
tapply(mtcars$mpg, mtcars$cyl, mean)
# mean mpg by cyl and gear
tapply(mtcars$mpg, list(cyl = mtcars$cyl, carb = mtcars$gear), mean)
```
### aggregate
```{r}
# same as above without NA
aggregate(mtcars$mpg, list(mtcars$cyl, mtcars$gear), mean)
# repeat over the data frame
aggregate(mtcars, list(mtcars$cyl, mtcars$gear), mean)
# using a formula
aggregate(cbind(mpg, wt) ~ cyl + gear, data = mtcars, mean)
# using the subset option
aggregate(cbind(mpg, wt) ~ cyl + gear, data = mtcars, mean, subset = carb > 4)
```
### by
```{r}
# summary of the data frame by group
by(mtcars, mtcars$am, summary)
```
## Packages
 * The `reshape2` package (a reboot of the `reshape` package) flexibly restructures and aggregates data using three functions `melt`, `acast` and `dcast`

 * The `plyr` package has tools for splitting, applying and combining data

## Exercise
* Create a 5x5 matrix from a vector `runif(25)`. Convert the matrix into a data frame using `as.data.frame`. Assign names X1, X2, ..., X5 to the columns.

* Using `which` and `is.na`, convert elements to NA that are larger than 0.9. 


* Consider the dataset `mtcars`. Define a function `perform.t = function(x) t.test(mpg ~ am, data = x)` which performs a t.test for testing mean difference in `mpg` between the `am=0` and `am=1` groups, given a subset of the data. Using this function and `by`, perform this t.test for each level of `cyl`. 

* Define another function that extracts the p-value of a t-test as `pv.t = function(x) x$p.v`. Using this function and `sapply`, extract p-values for the t-tests performed in the previous problem. 

* Repeat this for `perform2.t = function(x) t.test(wt ~ am, data = x)`. Combine the resulting vector of these p-values with the previously obtained vector using `cbind` with appropriate column names. 