Lab 1: Regression diagnostics
========================================================
```{r fig.width=4, fig.height=4}
data(women)
head(women)
lm.wo = lm(height ~ weight, data = women)
summary(lm.wo)
# height vs weight
plot(women$weight, women$height, xlab = "weight", ylab = "height")
lines(women$weight, fitted(lm.wo))
```
### Residual plots
```{r fig.width=4, fig.height=4}
# raw residuals
plot(fitted(lm.wo), resid(lm.wo), main = "Residuals vs fitted", xlab = "Fitted values", ylab = "Residuals")
abline(h = 0, lty = 2)
# standardized residuals
plot(fitted(lm.wo), rstandard(lm.wo), main = "Residuals vs Fitted", xlab = "Fitted values", ylab = "Standardized residuals")
abline(h = 0, lty = 2)
```
### Normal Q-Q plots
```{r fig.width=4, fig.height=4}
qqnorm(rstandard(lm.wo), main = "Normal Q-Q", ylab = "Standardized residuals")
abline(a = 0, b = 1)
```
Fit a quadratic curve. 
```{r fig.width=8, fig.height=3}
lm.wo2 = lm(height ~ poly(weight, 2), data = women)
summary(lm.wo2)
par(mfrow = c(1, 3))
# height vs weight
plot(women$weight, women$height, xlab = "weight", ylab = "height")
lines(women$weight, fitted(lm.wo2))
# standardized residuals
plot(fitted(lm.wo2), rstandard(lm.wo2), main = "Residuals vs fitted", xlab = "Fitted values", ylab = "Standardized residuals")
abline(h = 0, lty = 2)
# normal Q-Q plot
qqnorm(rstandard(lm.wo2), main = "Normal Q-Q", ylab = "Standardized residuals")
abline(a = 0, b = 1)
```
```{r echo = FALSE}
par(mfrow = c(1, 1))
```
### More examples
```{r echo = FALSE, fig.width=8, fig.height=2}
opar = par(no.readonly = T)
n = 100
par(mfrow = c(1, 4), mgp = c(1.5, .5, 0), mar = c(3,3,3,1) + .1)

# iid normal errors
x = 1:n
y = 1:n + rnorm(n)
lm1 = lm(y ~ x)
plot(x, y, main = "y vs x"); abline(lm1, lty = 2)
plot(fitted(lm1), rstandard(lm1), main = "Residuals vs fitted", xlab = "Fitted values", ylab = "Standardized residuals")
abline(h = 0, lty = 2)
qqnorm(rstandard(lm1), main = "Normal Q-Q", ylab = "Standardized residuals")
abline(a=0, b=1, lty = 2)
hist(rstandard(lm1), prob = T, main = "Histogram of residuals", xlab = "Standardized residuals")
curve(dnorm, add = T, col = "red")

# heavy tails
y = 1:n + rt(n, 2)
lm1 = lm(y ~ x)
plot(x, y, main = "y vs x"); abline(lm1, lty = 2)
plot(fitted(lm1), rstandard(lm1), main = "Residuals vs fitted", xlab = "Fitted values", ylab = "Standardized residuals")
abline(h = 0, lty = 2)
qqnorm(rstandard(lm1), main = "Normal Q-Q", ylab = "Standardized residuals")
abline(a=0, b=1, lty = 2)
legend("topleft", legend = "heavy tails", bty = "n", text.col = "red")
hist(rstandard(lm1), prob = T, main = "Histogram of residuals", xlab = "Standardized residuals")
curve(dnorm, add = T, col = "red")

# short tails
y = 1:n + runif(n)*n
lm1 = lm(y ~ x)
plot(x, y, main = "y vs x"); abline(lm1, lty = 2)
plot(fitted(lm1), rstandard(lm1), main = "Residuals vs fitted", xlab = "Fitted values", ylab = "Standardized residuals")
abline(h = 0, lty = 2)
qqnorm(rstandard(lm1), main = "Normal Q-Q", ylab = "Standardized residuals")
abline(a=0, b=1, lty = 2)
legend("topleft", legend = "short tails", bty = "n", text.col = "red")
hist(rstandard(lm1), prob = T, main = "Histogram of residuals", xlab = "Standardized residuals")
curve(dnorm, add = T, col = "red")

# nonlinear relationship
y = (1:n)^2 + rnorm(n)*100
lm1 = lm(y ~ x)
plot(x, y, main = "y vs x"); abline(lm1, lty = 2)
plot(fitted(lm1), rstandard(lm1), main = "Residuals vs fitted", xlab = "Fitted values", ylab = "Standardized residuals")
abline(h = 0, lty = 2)
legend("topleft", legend = "nonlinear relationship", bty = "n", text.col = "red")
qqnorm(rstandard(lm1), main = "Normal Q-Q", ylab = "Standardized residuals")
abline(a=0, b=1, lty = 2)
hist(rstandard(lm1), prob = T, main = "Histogram of residuals", xlab = "Standardized residuals")
curve(dnorm, add = T, col = "red")

# non-constant variance
y = 1:n + rnorm(n)*(1:n)
lm1 = lm(y ~ x)
plot(x, y, main = "y vs x"); abline(lm1, lty = 2)
plot(fitted(lm1), rstandard(lm1), main = "Residuals vs fitted", xlab = "Fitted values", ylab = "Standardized residuals")
abline(h = 0, lty = 2)
legend("topleft", legend = "non-constant variance", bty = "n", text.col = "red")
qqnorm(rstandard(lm1), main = "Normal Q-Q", ylab = "Standardized residuals")
abline(a=0, b=1, lty = 2)
hist(rstandard(lm1), prob = T, main = "Histogram of residuals", xlab = "Standardized residuals")
curve(dnorm, add = T, col = "red")

par(opar)
```

Exercise
--------------------------------------------------------
* Using the `cars` dataset in the `datasets` package, plot distance versus speed. Fit
a regression line and add the line to the existing plot. 

* Make a residuals plot and normal Q-Q plot for the residuals. 

* Fit and plot a quadratic curve. Also perform regression diagnostics as above. 

