Lab 2: Wilcoxon rank-sum test (a nonparametric hypothesis test)
========================================================
## Assumptions for independent two-sample t-test
* Two groups are independent 

* Each group comes from a normal distiribution (with equal/nonequal variance)

## Wilcoxon rank-sum test 
* Also called Mann–Whitney U test, Mann–Whitney–Wilcoxon (MWW), or Wilcoxon–Mann–Whitney test

* Nonparametric equivalent of the independent two-sample t-test 

* Assumes independence of two groups, but not normality

* Assumes the distributions of both groups are equal under the null hypothesis

* Robust to outliers

* Invariant under any rank-preserving transformation

* More efficient than the t-test on nonnormal distributions

* When normality holds, the rank-sum test is nearly as efficient as the t-test. 

```{r fig.width=4, fig.height=4}
Group = c(rep("A", 5), rep("B", 5)) # same as rep(c("A", "B"), rep(5, 2))
Outcome = c(16, 18, 20, 22, 24, 25, 28, 29, 33, 34)
boxplot(Outcome ~ Group)
t.test(Outcome ~ Group)
```
The t.test says two group means are significanly different. Suppose we have an outlier as follow. 
```{r fig.width=4, fig.height=4}
# what if no outlier
Outcome.o = Outcome
Outcome.o[10] = 60
boxplot(Outcome.o ~ Group)
t.test(Outcome.o ~ Group)
```
Now the t-test reaches the opposite conclusion. The rank-sum test is robust to outliers giving exactly the same test statistic and p-value.  
```{r}
wilcox.test(Outcome ~ Group)
wilcox.test(Outcome.o ~ Group)
```
In general, the rank-sum test is invariant under any rank-preserving transformation. 
```{r}
Outcome.log = log(Outcome)
wilcox.test(Outcome.log ~ Group)
Outcome.rank = rank(Outcome)
Outcome.rank
wilcox.test(Outcome.rank ~ Group)
```
You can use the `alternative="less"` or `alternative="greater"` option to specify a one-sided test.
## More nonparametric tests 
### Paired two-sample rank-sum test
Use `wilcox.test` with the option `paired = TRUE`.  
### One-way ANOVA Kruskal-Wallis test 
Use `kruskal.test` instead of `lm`. 

## Exercise 

A dataset from http://lib.stat.cmu.edu/DASL/: 
```
Datafile Name: Clouds

Story Names: Cloud Seeding

Description: Rainfall from Cloud-Seeding. The rainfall in acre-feet from 52 clouds 26 of which were chosen at random and seeded with silver nitrate.

Number of cases: 26

Variable Names:
Unseeded_Clouds: Amount of rainfall from unseeded clouds (in acre-feet)
Seeded_Clouds: Amount of rainfall from seeded clouds with silver nitrate (in acre-feet)
```

* For this dataset ('Clouds.txt'), make a side-by-side box plot for rainfall of the two groups. 

* Perform a two-sample t-test. One-sided or two-sided? 

* Perform a Wilcoxon rank-sum test. 

* Repeat the first two problems after taking the (natural) log of rainfall. 