Lab 3: Power analysis and sample size determination
========================================================
## Contingency table
        |$H_0$ is true	| $H_0$ is false |
:------ | :------ | :----- |
**Reject $H_0$** |	 Type I error | Correct |
      | False positive	| True positive |    
**Fail to reject $H_0$** |	Correct | Type II error |
    | True negative	| False negative |

## Components of power analysis

* Sample size $n$

* Effect size ES

* Significance level $\alpha$
= P(Type I error) = false positive rate

* Power $\beta$
= 1 $-$ P(Type II error) = true positive rate

One is expressed as a function of the other three. Therefore, given any three, we can calculate the fourth.

Power increases as sample size, effect size, or significance level increases.  

## Using the 'pwr' package
The `pwr` package has functions for power analysis. 
```{r eval = FALSE}
# (install and then) load the 'pwr' package
library(pwr)
```
### t-tests
```{r eval = FALSE}
# equal sample sizes
pwr.t.test(n = NULL, d = NULL, sig.level = 0.05, power = NULL, type = c("two.sample", "one.sample", "paired"), alternative = c("two.sided", "less","greater"))
```
```{r eval = FALSE}
# unequal sample sizes
pwr.t2n.test(n1 = NULL, n2 = NULL, d = NULL, sig.level = 0.05, power = NULL, alternative = c("two.sided", "less","greater"))
```

* `n` is the number of observations per sample.

* The `sig.level` default to 0.05 instead of `NULL`. Therefore, to calculate the significance level given the other three (`n`, `d`, and `power`), use the option `sig.level = NULL`.  
### Other tests 
http://www.statmethods.net/stats/power.html

## Cohen's recommendation for effect sizes
### t-tests 
$$d=\frac{|\mu_1-\mu_2|}{\sigma}$$ 	
with levels small = 0.2, medium = 0.5, large = 0.8. 
### Other tests, 
http://www.statmethods.net/stats/power.html

## Exercise 
* For a one-sided independent two-sample t-test, what is the sample size needed in each group to obtain a power of 0.8, when the significance level is set to be 0.01 and an effect size is 0.5 (medium)?

* With the sample size obtained above and with the significance level unchanged (0.01), what is the power when the effect size is 0.8 (large)? 
